Statistical analysis of whole-genome bisulfite-sequencing data

### 2014-12-02 Rmarkdown/wgbs_statistical_analysis.Rmd
Re-jigging of WGBS statistical analysis chapter

1. $\mathcal{I}^{ref}$: Defined by the set of methylation loci in the reference genome. This ignore all genetic variation between the sample and the reference.
2. $\mathcal{I}^{refFilter}$: Defined by filtering out problematic loci from $\mathcal{I}^{ref}$. A conservative approach that removes many sites of genetic variation between the sample and the reference as well as sites that do not display genetic variation between the sample and the reference. This approach cannot identify sample-specific methylation loci.
3. $\mathcal{I}^{Bis-SNP}$: Defined by calling genetic variants from the bisulfite-sequencing data using `Bis-SNP` \cite{Liu:2012ge}. Identifies sample-specific methylation loci and removes reference-specific methylation loci. This is the best approach if only bisulfite-sequencing data are available.
4. $\mathcal{I}^{WGS}$: Defined by identifying all methylation loci from _whole-genome sequencing_ (WGS) of the sample's genome. The gold standard. All methylation loci are defined with respect to the sample's genome. The only differences between $\mathcal{I}^{WGS}$ and $\mathcal{I}^{truth} are due to sequencing errors, incomplete sequencing coverage of the sample's genome and variant calling errors from the WGS data.

@@

Both `bsseq` and `BiSeq` use a binomial local likelihood smoother. In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by ($M_{i, j} | n_{i, j}$ \eqd Binom(n_{i, j}, B_{i, j})$. The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.

@@

\subsection{Experimental design}\label{sec:experimental_design}

@@

\subsection{DMCs}\label{sec:dmcs}

@@

\subsubsection{One-group experiments}

@@

\subsubsection{Multi-group experiments}\label{sec:two_group_dmcs}

@@

\subsection{DMRs}

@@

1. Usinb regions that are defined _a priori_, which are then tested for differential methylation. Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__) gene promoters (__CITE__) or predefined bins (e.g. 3bp tiles [Bing Ren] or 100bp tiles \citep{Park:2014ho}).
2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.

The former is simple to analyse but is limited in its ability to discover new differentially methylated regions. The latter offers the opportunity to identify previously unknown regions but valid statistical inference of these regions is __very challenging__.

__UP TO HERE__

@@

\subsection{DMRs}

__TODO: FDRs of region__

1. Use regions that are defined _a priori_, which are then tested for differential methylation. Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__) gene promoters (__CITE__) or predefined bins (e.g. 3bp tiles [Bing Ren] or 100bp tiles \citep{Park:2014ho}).
2. Use regions that are defined based on the data, which are then tested for differential methylation. Valid statistical inference of such regions is __very challenging__.

\subsubsection{Using _a priori_ regions}\label{sec:a_prior_regions}

The methylation level is summarised at the region level and then compared across groups. For example, the average of the $\beta$-values across the region may be compared between groups using a t-test.

__TODO: Describe the above mathematically__

This is qualitatively similar to identifying differentially methylated cytosines.

__PROS__

* Simple to implement
* Valid statistical properties

__CONS__

\subsubsection{Using data-driven regions}\label{sec:data_driven_regions}

__PROS__

__CONS__

* Loss of statistical properties in testing, e.g. FDR control
* Bloody hard

\section{Other downstream analyses}\label{sec:other_downstream_analyses}

Bisulfite-sequencing data are used to address a variety of questions apart from identifying differential methylation. Not all these analyses require statistical inference. For example, if we are interested in knowing whether the promoter of a particular gene is methylated in our sample then it may be sufficient to simply plot the level of methylation at each CpG in the region.

However, once we move to experiments with multiple samples, multiple regions and more complex questions then statistical methods are generally required to distinguish the signal from the noise. In this section I briefly describe other

__TODO: Any other important downstream analyses worth mentioning?__

@@

\subsection{General computational considerations}

- Using integers instead of floats/doubles

@@

* Autocorrelation or correlation?
* Discuss distribution of coverage? Perhaps in context of sequencing bias?
* See https://github.com/brentp/clustermodel/README.md for a nice, brief summary of the main DMR calling approaches.

>>>>

1. $\mathcal{I}^{ref}$: Defined by the set of methylation loci in the reference genome. This ignore all genetic variation between the sample and the reference.
2. $\mathcal{I}^{refFilter}$: Defined by filtering out problematic loci from $\mathcal{I}^{ref}$. A conservative approach that removes many sites of genetic variation between the sample and the reference as well as sites that do not display genetic variation between the sample and the reference. This approach cannot identify sample-specific methylation loci.
3. $\mathcal{I}^{Bis-SNP}$: Defined by calling genetic variants from the bisulfite-sequencing data using `Bis-SNP` \cite{Liu:2012ge}. Identifies sample-specific methylation loci and removes reference-specific methylation loci. This is the best approach if only bisulfite-sequencing data are available.
4. $\mathcal{I}^{WGS}$: Defined by identifying all methylation loci from _whole-genome sequencing_ (WGS) of the sample's genome. The gold standard. All methylation loci are defined with respect to the sample's genome. The only differences between $\mathcal{I}^{WGS}$ and $\mathcal{I}^{truth}$ are due to sequencing errors, incomplete sequencing coverage of the sample's genome and variant calling errors from the WGS data.

@@

Both `bsseq` and `BiSeq` use a binomial local likelihood smoother. In each case this smoother was chosen because `BSmooth` and `BiSeq` model the number of methylated reads at the $i^{th}$ locus in the $j^{th}$ sample by ($M_{i, j} | n_{i, j} \eqd Binom(n_{i, j}, B_{i, j})$. The smoothing is 'local' to exploit co-methylation, which is a local phenomenon.

@@

\subsubsection{Experimental design}\label{sec:experimental_design}

@@

\subsubsection{DMCs}\label{sec:dmcs}

@@

\paragraph{One-group experiments}\label{sec:one_group_dmcs}

@@

\paragraph{Multi-group experiments}\label{sec:two_group_dmcs}

@@

\subsubsection{DMRs}

A differentially methylated region is a region of the genome where there are multiple cytosines with evidence of differential methylation. Importantly, not all cytosines in the region must be genome-wide significant DMCs. Rather, the idea is that a DMR might capture a weaker but broader difference in methylation. Such a signal may well have a clearer biological interpretation and, indeed, might be expected such as in differences in methylation at a gene promoter.

__UP TO HERE: Take care of TODOs and then move onto other downstream analyses.__

@@

1. Using regions that are defined _a priori_, which are then tested for differential methylation. Such regions might be CpG islands (__CITE__), MspI restriction fragments (__CITE__), gene promoters (__CITE__) or predefined bins \citep[e.g., 100 bp bins]{Park:2014ho}.
2. Using data-driven regions, e.g., defined from an analysis of DMCs, which are then tested for differential methylation.

The former is much simpler to analyse but is limited in its ability to discover new differentially methylated regions. The latter offers the opportunity to identify previously unknown regions but valid statistical inference of these regions is very challenging.

__TODO: Further cons of a priori regions__

__TODO: Further pros of data-driven regions__

The most common type of algorithm for identifying data-driven DMRs is based on detecting clusters of DMCs (__CITE: e.g., Lister, Hansen, DSS?, Berman?, etc.). This initial identification of DMCs will typically use a relaxed statistical significance threshold. For example, \cite{Hansen:2011gu} use all CpGs with a P-value in the lowest $5\%$ and declare putative DMRs to be contiguous runs of such CpGs that are within a given distance of one another and with "all differences in the same direction". This latter requirement is what I call the _consistency_ of the DMR (see Section \ref{sec:consistency_of_dmrs}).

[^a_prior]: In contrast, DMR analyses using _a priori_ regions do not need to first test for DMCs, although this single-locus analysis may still be performed.

__TODO: Other data-driven DMR detection strategies__
- bump-hunting
- Mixed effect model
- GEE, i.e., `Aclust` distinguishes itself by using data-driven regions to identify DMRs but without first scanning the genome for DMCs.
- probe aggregation, e.g., `comb-P`

Finally, __TODO: Do these methods actually test for differential methylation at these regions? If they do, then a test of a DMR requires a statistic that summarises the evidence that the region is differentially methylated. What about FDRs?__

For each sample, the region-wise methylation level is some summary of the methylation level at each loci in that region. For example, the weighted average of $\beta$-values for all loci in the region, where the weights are proportional to the sequencing coverage. These averages are then compared between samples.

\paragraph{One-group experiments}\label{sec:one_group_dmrs}

For a region containing $c$ methylation loci, construct the $n \times 2 \times c$ contingency table, i.e., each 'slice' of the table is a $n \times 2$ table containing the data for a single locus. Simply aggregating the counts across loci and testing the resulting $n \times 2$ table would be incorrect, since such an aggregating could introduce spurious differences in the proportions (__CITE: http://www.biostathandbook.com/cmh.html__). Instead, we could fit the log-linear model with `sample`, `meth_type` and `locus` as the set of factors. The saturated model is

\begin{equation*}
\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus} + \lambda_{a, b, c}^{meth_state, sample, locus}
\end{equation*}

We would first need to test whether there is a significant three-way interaction between `meth_type`, `sample` and `locus` by computing the deviance of the model without the three-way interaction

\begin{equation*}
\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
\end{equation*}

On the one hand, if there is no three-way interaction, then we can test for an interaction between `meth_type` and `sample`, i.e., test for a difference in methylation level across the loci, by comparing the model

\begin{equation*}
\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, b}^{meth_state, sample} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
\end{equation*}

to the model

\begin{equation*}
\log(F_{a, b, c}) = \lambda_{0} + \lambda_{a}^{meth_state} + \lambda_{b}^{sample} + \lambda_{c}^{locus} + \lambda_{a, c}^{meth_state, locus} + \lambda_{b, c}^{sample, locus}
\end{equation*}

This test is analogous to the generalised Cochran-Mantel-Haenszel test \citep{Landis:1978vk,MANTEL:1978iv,Birch:1965uj} of the $n \times 2 \times c$ table \citep[p231]{agresti1990categorical}. The advantage of the Cochran-Mantel-Haenszel is its applicability for sparse datasets for which the asymptotic theory does not hold for the test based on the log-linear regression  \citep[p232]{agresti1990categorical}.

On the other hand, if there is a significant three-way interaction between `meth_type`, `sample` and `locus` then this requires more careful modelling to determine the nature of this interaction and whether the region is a consistent DMR.

\paragraph{Multi-group experiments}\label{sec:multi_group_dmrs}

__TODO__

\paragraph{Consistency of DMRs}\label{sec:consistency_of_dmrs}

While not often explicitly stated, it is desirable that a DMR is _consistent_. The definition of consistency is best illustrated by a pathological example describing a region that is inconsistent. Suppose we have a one-group experiment with five samples[^one_by_five] and are studying a region with five CpGs, all of which are DMCs. However, the first CpG is differentially methylated in the first sample, the second CpG is differentially methylated in the second sample, etc. While the entire region is 'differentially methylated', the difference is not observed consistently in any one sample. A consistent region is therefore one where the differences are 'in the same direction'.

[^one_by_five]: Equivalently, a five group experiment without replicates.

Consistent DMRs are most easily identified in two-group experiments, where we can use the sign of the test statistic to determine the 'direction' of the signal. This is exactly what is done by, for example, \cite{Hansen:2011gu}.

In one-group experiments, the idea of consistency is similar to testing the homogeneity of odds ratios for all $n \times 2$ contigency tables in the region. However, __TODO. What about a region that is consistent but the odds ratios increases across the region__

@@

@@

@@

* Discuss distribution of coverage? Perhaps in context of sequencing bias?
